---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
####   1. Read the base zip file "activity.zip" that contains the data
#####     This first line will likely take a few seconds. Be patient!

```{r, echo = TRUE, results = 'markup', warning = TRUE, message = TRUE}

dataInput_file <- "activity.zip"
if (!file.exists(dataInput_file)) {
  data_URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(data_URL, destfile = dataInput_file)
  unzip (zipfile = dataInput_file)
  
}
activityData <- read.csv('activity.csv')

summary(activityData)


```
##### 2. Load the libraries 

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)
library(Hmisc)
```
##### 3. Transforming Interval Data

```{r}
#activityData$interval <- strptime(gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", activityData$interval), format='%H:%M')
```

-----

## What is mean total number of steps taken per day?


```{r}
stepsByDay <- tapply(activityData$steps, activityData$date, sum, na.rm=TRUE)
```


##### 1. Histogram of the total number of steps taken each day

```{r}
qplot(stepsByDay, xlab='Total steps per day', ylab='Frequency using binwith 500', binwidth=500)
```



##### 2. Mean and median total number of steps taken per day

```{r}
stepsByDayMean <- mean(stepsByDay)
stepsByDayMedian <- median(stepsByDay)
```
* Mean: `r stepsByDayMean` 
* Median: `r stepsByDayMedian`

-----

## What is the average daily activity pattern?


```{r}
averageStepsPerTimeBlock <- aggregate(x=list(meanSteps=activityData$steps), by=list(interval=activityData$interval), FUN=mean, na.rm=TRUE)
```

##### 1. Time series plot

```{r}
ggplot(data=averageStepsPerTimeBlock, aes(x=interval, y=meanSteps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken") 
```
##### 2. The 5-minute interval on average across all the days in the dataset that contains the maximum number of steps?

```{r}
mostSteps <- which.max(averageStepsPerTimeBlock$meanSteps)
timeMostSteps <-  gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", averageStepsPerTimeBlock[mostSteps,'interval'])
```

* Most Steps at: `r timeMostSteps`

----


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
